---
title: "Sounds of San Anto: Music Venues & Neighborhood Change, 1970–2010"
author: "Erick Lima | CEDISH / UTSA"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    self_contained: true
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = "/Users/erick/CEDISH/San_Anto")
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

library(tidyverse)
library(readxl)
library(sf)
library(tigris)
library(leaflet)
library(leaflet.extras)
library(htmlwidgets)
library(htmltools)
library(RColorBrewer)
library(tidycensus)

options(tigris_use_cache = TRUE)
```

---

# Introduction

This notebook builds a layered interactive HTML map telling a gentrification
story: how concert venue geography in San Antonio correlates with neighborhood
income and rent changes from 1980–2000, with race, housing tenure, population
shift, venue churn, event volume collapse, and genre displacement as additional
analytical layers. The map integrates 312 geocoded music venues (1970–2010),
NHGIS census data, and tidycensus decennial estimates for Bexar County.

---

# Step 1 — Load and Clean Events

```{r load-events}
events_raw <- read_excel("san_anto_1.xlsx", sheet = "Ven_Ev_Ar_Address")

events <- events_raw %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  mutate(Longitude = ifelse(Longitude > 0, -Longitude, Longitude)) %>%
  mutate(
    Year   = as.integer(Year),
    decade = floor(Year / 10) * 10
  )

cat("Geocoded events retained:", nrow(events), "of", nrow(events_raw), "\n")
```

```{r load-genres}
artists_genre <- read_excel("san_anto_1.xlsx", sheet = "artistsGenre") %>%
  distinct(Name, .keep_all = TRUE)

macro_genre_crosswalk <- function(genre) {
  case_when(
    str_detect(genre, regex(
      "country|outlaw|heartland|western swing|honky|bluegrass|rockabilly|
       nashville|americana|texas country|classic country|cowboy|cowpunk",
      ignore_case = TRUE, comments = TRUE)) ~ "Country",

    str_detect(genre, regex(
      "tejano|conjunto|norteño|latin|salsa|cumbia|mariachi|ranchera|banda|
       chicano|mexican|hispanico|tropical|bolero|corrido",
      ignore_case = TRUE, comments = TRUE)) ~ "Tejano/Latin",

    str_detect(genre, regex(
      "classical|symphony|orchestra|opera|chamber|choral|baroque|
       contemporary classical|avant.garde classical",
      ignore_case = TRUE, comments = TRUE)) ~ "Classical",

    str_detect(genre, regex(
      "rock|punk|grunge|alternative|indie|garage|psychedelic|progressive rock|
       hard rock|post.punk|new wave|surf rock",
      ignore_case = TRUE, comments = TRUE)) ~ "Rock",

    str_detect(genre, regex(
      "jazz|bebop|swing|fusion|big band|smooth jazz|cool jazz|free jazz",
      ignore_case = TRUE, comments = TRUE)) ~ "Jazz",

    str_detect(genre, regex(
      "blues|r&b|rhythm|soul|funk|gospel|motown|doo.wop",
      ignore_case = TRUE, comments = TRUE)) ~ "Blues/R&B",

    str_detect(genre, regex(
      "metal|heavy metal|thrash|doom|death metal|black metal|glam metal",
      ignore_case = TRUE, comments = TRUE)) ~ "Metal",

    str_detect(genre, regex(
      "pop|adult contemporary|top 40|dance.pop|bubblegum",
      ignore_case = TRUE, comments = TRUE)) ~ "Pop",

    str_detect(genre, regex(
      "electronic|techno|house|edm|synth|ambient|trance|drum.and.bass|
       industrial|electronica",
      ignore_case = TRUE, comments = TRUE)) ~ "Electronic",

    is.na(genre) ~ "Unknown/Other",
    TRUE          ~ "Other"
  )
}

artists_genre <- artists_genre %>%
  mutate(macro_genre = macro_genre_crosswalk(Genres))

events <- events %>%
  left_join(artists_genre %>% select(Name, Genres, macro_genre),
            by = c("Artist" = "Name")) %>%
  mutate(macro_genre = replace_na(macro_genre, "Unknown/Other"))

cat("Events missing genre data:", sum(events$macro_genre == "Unknown/Other"), "\n")
```

```{r events-sf}
events_sf <- events %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
```

---

# Step 2 — Census Tract Geometries

```{r load-tracts}
tracts_bexar <- tracts(state = "TX", county = "Bexar", year = 2000, cb = TRUE) %>%
  st_transform(crs = 4326) %>%
  mutate(GEOID = paste0(STATE, COUNTY, TRACT))

cat("Bexar County tracts loaded:", nrow(tracts_bexar), "\n")
```

---

# Step 3 — NHGIS Economic Data

NHGIS GISJOIN formats differ by year (1980/1990 = 13-char, 2000 = 15-char).
We convert all to the standard 11-char GEOID used by tigris.

```{r load-nhgis}
rent_1980 <- read_csv("nhgis0001_csv/nhgis0001_ds107_1980_tract.csv") %>%
  filter(COUNTY == "Bexar") %>%
  transmute(GISJOIN, rent_1980 = DFK001)

rent_1990 <- read_csv("nhgis0001_csv/nhgis0001_ds123_1990_tract.csv") %>%
  filter(COUNTY == "Bexar") %>%
  transmute(GISJOIN, rent_1990 = EYU001)

rent_2000 <- read_csv("nhgis0001_csv/nhgis0001_ds151_2000_tract.csv") %>%
  filter(COUNTY == "Bexar") %>%
  transmute(GISJOIN, rent_2000 = GBO001)

income_ts <- read_csv("nhgis0001_csv/nhgis0001_ts_nominal_tract.csv") %>%
  filter(COUNTY == "Bexar County") %>%
  select(GJOIN1980, GJOIN1990, GJOIN2000, B79AA1980, B79AA1990, B79AA2000)
```

```{r gisjoin-helpers}
gisjoin_to_geoid <- function(gisjoin, tract_digits) {
  state_part  <- substr(gisjoin, 2, 3)
  county_part <- substr(gisjoin, 5, 7)
  tract_raw   <- substr(gisjoin, 9, 8 + tract_digits)
  tract_part  <- str_pad(tract_raw, 6, side = "right", pad = "0")
  paste0(state_part, county_part, tract_part)
}

rent_1980 <- rent_1980 %>% mutate(GEOID = gisjoin_to_geoid(GISJOIN, 4))
rent_1990 <- rent_1990 %>% mutate(GEOID = gisjoin_to_geoid(GISJOIN, 4))
rent_2000 <- rent_2000 %>% mutate(GEOID = gisjoin_to_geoid(GISJOIN, 6))

income_wide <- income_ts %>%
  mutate(GEOID = gisjoin_to_geoid(GJOIN2000, 6)) %>%
  rename(income_1980 = B79AA1980, income_1990 = B79AA1990, income_2000 = B79AA2000)
```

```{r join-tract-data}
dedup_geoid <- function(df, val_col) {
  df %>%
    group_by(GEOID) %>%
    summarise(across(all_of(val_col), ~ first(na.omit(.x))), .groups = "drop")
}

econ_data <- dedup_geoid(rent_1980, "rent_1980") %>%
  full_join(dedup_geoid(rent_1990, "rent_1990"), by = "GEOID") %>%
  full_join(dedup_geoid(rent_2000, "rent_2000"), by = "GEOID") %>%
  full_join(
    income_wide %>%
      group_by(GEOID) %>%
      summarise(across(c(income_1980, income_1990, income_2000),
                       ~ first(na.omit(.x))), .groups = "drop"),
    by = "GEOID"
  ) %>%
  mutate(rent_change_80_00 = rent_2000 - rent_1980)

tracts_econ <- tracts_bexar %>%
  left_join(econ_data, by = "GEOID")

cat("Tracts with rent_1980:", sum(!is.na(tracts_econ$rent_1980)), "\n")
cat("Tracts with rent_2000:", sum(!is.na(tracts_econ$rent_2000)), "\n")
```

---

# Step 4 — Race, Ethnicity, Housing Tenure & Displacement Risk

We extend the tidycensus pull to capture Hispanic %, Black %, and
renter-occupied %. Black population gives a fuller picture of the racialized
geography of displacement (especially the eastside); housing tenure flags
which tracts are most vulnerable (high renter share = more exposure to rent
increases). We then build a **Displacement Risk Index** (0–100) by combining
rent change 1980–2000 and renter-occupied % — both min-max normalized and
averaged. This composite captures both the economic pressure and the
residential vulnerability in each tract without requiring unavailable
historical population data.

```{r census-2000-extended}
# ── Single 2000 SF1 pull: race + housing tenure ───────────────────────────────
# P004001 = Total pop  |  P004002 = Hispanic/Latino  |  P003003 = Black alone
# H004001 = Total occupied units  |  H004002 = Owner-occ  |  H004003 = Renter-occ
census_2000 <- get_decennial(
  geography = "tract",
  state     = "TX",
  county    = "Bexar",
  year      = 2000,
  variables = c(
    total_pop   = "P004001",
    hisp_pop    = "P004002",
    black_pop   = "P003003",
    owner_occ   = "H004002",
    renter_occ  = "H004003"
  ),
  output = "wide"
) %>%
  mutate(
    pct_hispanic = ifelse(total_pop > 0,
                          round(hisp_pop   / total_pop * 100, 1), NA_real_),
    pct_black    = ifelse(total_pop > 0,
                          round(black_pop  / total_pop * 100, 1), NA_real_),
    pct_renter   = ifelse((owner_occ + renter_occ) > 0,
                          round(renter_occ / (owner_occ + renter_occ) * 100, 1),
                          NA_real_)
  ) %>%
  select(GEOID, total_pop, hisp_pop, pct_hispanic,
         black_pop, pct_black, owner_occ, renter_occ, pct_renter)
```

```{r displacement-risk}
# ── Displacement Risk Index ───────────────────────────────────────────────────
# tidycensus get_decennial() only supports 2000/2010/2020, so 1990 population
# is unavailable. Instead we build a composite Displacement Risk Index (0–100)
# from two variables already in hand:
#   - Rent pressure:  rent_change_80_00 (economic stress on the tract)
#   - Renter exposure: pct_renter       (share of households vulnerable to it)
# Each variable is min-max normalized to [0,1], then averaged.
# High scores = neighborhoods where rent rose fast AND most residents rented
# = highest displacement risk.

normalize_01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0, length(x)))
  (x - rng[1]) / diff(rng)
}

displacement_risk <- census_2000 %>%
  select(GEOID, pop_2000 = total_pop) %>%
  left_join(econ_data %>% select(GEOID, rent_change_80_00), by = "GEOID") %>%
  left_join(census_2000 %>% select(GEOID, pct_renter),      by = "GEOID") %>%
  mutate(
    norm_rent_change = normalize_01(replace_na(rent_change_80_00, 0)),
    norm_renter      = normalize_01(replace_na(pct_renter, 0)),
    displacement_risk_idx = round((norm_rent_change + norm_renter) / 2 * 100, 1)
  ) %>%
  select(GEOID, pop_2000, displacement_risk_idx)

cat("Tracts with displacement risk index:",
    sum(!is.na(displacement_risk$displacement_risk_idx)), "\n")
```

```{r build-tracts-full}
# Join all census data to tract geometries
tracts_full <- tracts_econ %>%
  left_join(census_2000 %>% select(GEOID, pct_hispanic, pct_black, pct_renter,
                                    total_pop, hisp_pop, black_pop),
            by = "GEOID") %>%
  left_join(displacement_risk %>% select(GEOID, pop_2000, displacement_risk_idx),
            by = "GEOID")

cat("Tracts with race data:",    sum(!is.na(tracts_full$pct_hispanic)), "\n")
cat("Tracts with Black % data:", sum(!is.na(tracts_full$pct_black)),    "\n")
cat("Tracts with renter data:",  sum(!is.na(tracts_full$pct_renter)),   "\n")
cat("Tracts with risk index:",   sum(!is.na(tracts_full$displacement_risk_idx)), "\n")
```

---

# Step 5 — Load Venues and Build Popups

```{r load-venues}
venues_raw <- read_excel("san_anto_1.xlsx", sheet = "venues_with_coordinates") %>%
  filter(needs_geo == FALSE) %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude  = as.numeric(Latitude),
    Zip       = as.character(Zip),
    Longitude = ifelse(Longitude > 0, -Longitude, Longitude)
  ) %>%
  filter(!is.na(Longitude), !is.na(Latitude))

cat("Geocoded venues:", nrow(venues_raw), "\n")
```

```{r venue-summaries}
top_genres_for_venue <- function(genres) {
  tbl <- sort(table(genres[!is.na(genres) & genres != "Unknown/Other"]),
              decreasing = TRUE)
  if (length(tbl) == 0) return("Unknown")
  paste(names(head(tbl, 3)), collapse = ", ")
}

venue_summary <- events %>%
  group_by(Venue) %>%
  summarise(
    n_events   = n(),
    min_year   = min(Year, na.rm = TRUE),
    max_year   = max(Year, na.rm = TRUE),
    top_genres = top_genres_for_venue(macro_genre),
    .groups    = "drop"
  )

venues_enriched <- venues_raw %>%
  left_join(venue_summary, by = c("Venue_Name" = "Venue")) %>%
  mutate(
    n_events   = replace_na(n_events, 0),
    top_genres = replace_na(top_genres, "Unknown")
  )
```

```{r venue-sf}
venues_sf <- venues_enriched %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

venues_sf <- st_join(
  venues_sf,
  tracts_full %>% select(GEOID, rent_1980, rent_2000, rent_change_80_00,
                          income_2000, pct_hispanic, pct_black, pct_renter),
  join = st_within
) %>%
  distinct(Venue_Name, .keep_all = TRUE)

venues_sf <- venues_sf %>%
  mutate(
    popup_html = paste0(
      "<div style='font-family:sans-serif;font-size:13px;max-width:260px'>",
      "<b style='font-size:14px'>", Venue_Name, "</b><br>",
      "<span style='color:#555'>", Address, " · ", Zip, "</span>",
      "<hr style='margin:4px 0'>",
      "Active: <b>", ifelse(n_events > 0,
                            paste0(min_year, "–", max_year), "Unknown"),
      "</b> · Events: <b>", n_events, "</b><br>",
      "Top genres: <i>", top_genres, "</i>",
      "<hr style='margin:4px 0'>",
      "Nbhd rent 2000: <b>",
        ifelse(!is.na(rent_2000), paste0("$", format(rent_2000, big.mark=",")), "N/A"),
        "</b> · Rent Δ: <b>",
        ifelse(!is.na(rent_change_80_00),
               paste0("$", format(round(rent_change_80_00), big.mark=",")), "N/A"),
        "</b><br>",
      "% Hispanic: <b>",
        ifelse(!is.na(pct_hispanic), paste0(pct_hispanic, "%"), "N/A"),
        "</b> · % Black: <b>",
        ifelse(!is.na(pct_black), paste0(pct_black, "%"), "N/A"),
        "</b><br>",
      "% Renter-occupied: <b>",
        ifelse(!is.na(pct_renter), paste0(pct_renter, "%"), "N/A"),
        "</b>",
      "</div>"
    )
  )
```

```{r decade-venue-layers}
decade_activity <- events %>%
  filter(!is.na(decade)) %>%
  group_by(Venue, decade) %>%
  summarise(n_decade = n(), .groups = "drop")

make_decade_layer <- function(dec) {
  active <- decade_activity %>% filter(decade == dec) %>% pull(Venue)
  venues_sf %>% filter(Venue_Name %in% active)
}

venues_1970s <- make_decade_layer(1970)
venues_1980s <- make_decade_layer(1980)
venues_1990s <- make_decade_layer(1990)
venues_2000s <- make_decade_layer(2000)

cat("Venues active — 1970s:", nrow(venues_1970s),
    "| 1980s:", nrow(venues_1980s),
    "| 1990s:", nrow(venues_1990s),
    "| 2000s:", nrow(venues_2000s), "\n")
```

---

# Step 5b — Venue Churn Analysis

Venue churn reveals the direct cultural cost of gentrification: which
neighborhoods lost music venues between the 1980s and 2000s even as rents rose?
We identify venues active in the 1980s but gone by the 2000s (lost), and new
venues that appeared in the 2000s (gained), then aggregate both by census tract.

```{r venue-churn}
venues_1980s_names <- decade_activity %>% filter(decade == 1980) %>% pull(Venue)
venues_2000s_names <- decade_activity %>% filter(decade == 2000) %>% pull(Venue)

lost_names   <- setdiff(venues_1980s_names, venues_2000s_names)
gained_names <- setdiff(venues_2000s_names, venues_1980s_names)

venues_lost_sf   <- venues_sf %>% filter(Venue_Name %in% lost_names)
venues_gained_sf <- venues_sf %>% filter(Venue_Name %in% gained_names)

# Popup for churn markers
venues_lost_sf <- venues_lost_sf %>%
  mutate(popup_html = paste0(
    "<div style='font-family:sans-serif;font-size:13px;max-width:240px'>",
    "<b style='color:#c0392b'>&#x274C; LOST: ", Venue_Name, "</b><br>",
    "<span style='color:#555'>", Address, "</span><br>",
    "Active 1980s · not present by 2000s<br>",
    "Rent change in tract: <b>",
      ifelse(!is.na(rent_change_80_00),
             paste0("$", format(round(rent_change_80_00), big.mark=",")), "N/A"),
    "</b></div>"
  ))

venues_gained_sf <- venues_gained_sf %>%
  mutate(popup_html = paste0(
    "<div style='font-family:sans-serif;font-size:13px;max-width:240px'>",
    "<b style='color:#27ae60'>&#x2705; NEW: ", Venue_Name, "</b><br>",
    "<span style='color:#555'>", Address, "</span><br>",
    "New in 2000s · not present in 1980s<br>",
    "Rent change in tract: <b>",
      ifelse(!is.na(rent_change_80_00),
             paste0("$", format(round(rent_change_80_00), big.mark=",")), "N/A"),
    "</b></div>"
  ))

# Tract-level churn counts (join back to tracts_full)
lost_by_tract <- venues_lost_sf %>%
  st_drop_geometry() %>%
  filter(!is.na(GEOID)) %>%
  count(GEOID, name = "venues_lost")

gained_by_tract <- venues_gained_sf %>%
  st_drop_geometry() %>%
  filter(!is.na(GEOID)) %>%
  count(GEOID, name = "venues_gained")

venue_churn_tract <- lost_by_tract %>%
  full_join(gained_by_tract, by = "GEOID") %>%
  mutate(
    venues_lost   = replace_na(venues_lost,   0),
    venues_gained = replace_na(venues_gained, 0),
    net_venue_change = venues_gained - venues_lost
  )

tracts_full <- tracts_full %>%
  left_join(venue_churn_tract, by = "GEOID") %>%
  mutate(
    venues_lost      = replace_na(venues_lost,      0),
    venues_gained    = replace_na(venues_gained,    0),
    net_venue_change = replace_na(net_venue_change, 0)
  )

cat("Venues lost (1980s → 2000s):", length(lost_names), "\n")
cat("Venues gained (new in 2000s):", length(gained_names), "\n")
cat("Tracts that lost at least 1 venue:", sum(tracts_full$venues_lost > 0), "\n")
```

---

# Step 5c — Event Volume Collapse by Tract

Even where venues survived, total event volume could collapse — a subtler form
of cultural erosion. We spatially assign each event to a census tract and
compare total event counts between the 1980s and 2000s per tract.

```{r event-volume}
events_tract <- st_join(
  events_sf %>% filter(decade %in% c(1980, 2000)),
  tracts_full %>% select(GEOID),
  join = st_within
) %>%
  st_drop_geometry() %>%
  filter(!is.na(GEOID))

event_volume <- events_tract %>%
  group_by(GEOID, decade) %>%
  summarise(n_events = n(), .groups = "drop") %>%
  pivot_wider(names_from = decade, values_from = n_events,
              names_prefix = "events_", values_fill = 0) %>%
  rename(events_1980s = events_1980, events_2000s = events_2000) %>%
  mutate(
    event_change_abs = events_2000s - events_1980s,
    event_change_pct = ifelse(
      events_1980s > 0,
      round((events_2000s - events_1980s) / events_1980s * 100, 1),
      NA_real_
    )
  )

tracts_full <- tracts_full %>%
  left_join(event_volume, by = "GEOID") %>%
  mutate(
    events_1980s     = replace_na(events_1980s, 0),
    events_2000s     = replace_na(events_2000s, 0),
    event_change_abs = replace_na(event_change_abs, 0)
  )

cat("Tracts with event volume data:", sum(!is.na(tracts_full$event_change_pct)), "\n")
cat("Tracts where event volume fell:", sum(tracts_full$event_change_abs < 0, na.rm=TRUE), "\n")
```

---

# Step 5d — Genre Displacement by Tract

We compute the dominant genre per tract for the 1980s and 2000s separately,
then flag tracts where the top genre shifted — a proxy for cultural displacement
alongside demographic change.

```{r genre-displacement}
dominant_genre_by_decade <- events_tract %>%
  filter(macro_genre != "Unknown/Other") %>%
  group_by(GEOID, decade, macro_genre) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(GEOID, decade) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(GEOID, decade, dominant_genre = macro_genre) %>%
  pivot_wider(names_from = decade, values_from = dominant_genre,
              names_prefix = "genre_")

# Rename columns if they exist
if ("genre_1980" %in% names(dominant_genre_by_decade)) {
  dominant_genre_by_decade <- dominant_genre_by_decade %>%
    rename(genre_1980s = genre_1980, genre_2000s = genre_2000)
}

dominant_genre_by_decade <- dominant_genre_by_decade %>%
  mutate(genre_shifted = !is.na(genre_1980s) & !is.na(genre_2000s) &
                          genre_1980s != genre_2000s)

tracts_full <- tracts_full %>%
  left_join(dominant_genre_by_decade, by = "GEOID")

cat("Tracts with genre data in both decades:",
    sum(!is.na(tracts_full$genre_1980s) & !is.na(tracts_full$genre_2000s)), "\n")
cat("Tracts where dominant genre shifted:",
    sum(tracts_full$genre_shifted, na.rm = TRUE), "\n")
```

---

# Step 6 — Build the Layered Leaflet Map

The map has 11 choropleth overlays and 7 venue point layers, all independently
toggleable. Use the layer control (top-right, collapsed by default) to explore.

```{r color-palettes}
# ── Rent ──────────────────────────────────────────────────────────────────────
pal_rent_1980   <- colorBin("Blues", domain = tracts_full$rent_1980,
                              bins = 6, na.color = "#cccccc")
pal_rent_2000   <- colorBin("Blues", domain = tracts_full$rent_2000,
                              bins = 6, na.color = "#cccccc")
pal_rent_change <- colorBin(
  palette  = c("#2166ac","#92c5de","#f7f7f7","#f4a582","#d6604d","#b2182b"),
  domain   = tracts_full$rent_change_80_00,
  bins     = 7, na.color = "#cccccc"
)

# ── Race / ethnicity ──────────────────────────────────────────────────────────
pal_hispanic <- colorBin("Greens",  domain = tracts_full$pct_hispanic,
                          bins = 6, na.color = "#cccccc")
pal_black    <- colorBin("Purples", domain = tracts_full$pct_black,
                          bins = 6, na.color = "#cccccc")

# ── Housing tenure ────────────────────────────────────────────────────────────
pal_renter <- colorBin("Oranges", domain = tracts_full$pct_renter,
                        bins = 6, na.color = "#cccccc")

# ── Displacement Risk Index (yellow → red) ────────────────────────────────────
pal_risk <- colorBin("YlOrRd", domain = tracts_full$displacement_risk_idx,
                      bins = 6, na.color = "#cccccc")

# ── Venue churn (net) — red = net loss, green = net gain ─────────────────────
churn_range <- range(tracts_full$net_venue_change, na.rm = TRUE)
pal_venue_churn <- colorBin(
  palette  = c("#b2182b","#ef8a62","#fddbc7","#f7f7f7","#a1d99b","#41ab5d","#006d2c"),
  domain   = tracts_full$net_venue_change,
  bins     = 7, na.color = "#eeeeee"
)

# ── Event volume change (diverging) ──────────────────────────────────────────
pal_event_change <- colorBin(
  palette  = c("#b2182b","#ef8a62","#fddbc7","#f7f7f7","#d1e5f0","#67a9cf","#2166ac"),
  domain   = tracts_full$event_change_pct,
  bins     = 7, na.color = "#cccccc"
)

# ── Genre categories ──────────────────────────────────────────────────────────
genre_levels <- c("Country","Tejano/Latin","Classical","Rock","Jazz",
                   "Blues/R&B","Metal","Pop","Electronic","Other","Unknown/Other")
pal_genre <- colorFactor(
  palette  = c(brewer.pal(8, "Dark2"), "#aaaaaa", "#888888", "#555555"),
  domain   = genre_levels,
  na.color = "#555555"
)

# Dominant genre palette (same colors, applied to tract polygons)
genre_palette_vec <- setNames(
  c(brewer.pal(8, "Dark2"), "#aaaaaa", "#888888", "#555555"),
  genre_levels
)
pal_dom_genre <- colorFactor(
  palette = genre_palette_vec,
  domain  = genre_levels,
  na.color = "#eeeeee"
)
```

```{r build-map}
title_html <- tags$div(
  style = paste(
    "position:fixed; top:10px; left:50%; transform:translateX(-50%);",
    "background:rgba(255,255,255,0.93); padding:8px 18px; border-radius:6px;",
    "box-shadow:0 2px 8px rgba(0,0,0,.25); z-index:900;",
    "font-family:sans-serif; font-size:14px; font-weight:700;",
    "color:#222; text-align:center; max-width:90vw;"
  ),
  "Sounds of San Anto: Music Venues & Neighborhood Change in San Antonio, 1970–2010"
)

note_html <- tags$div(
  style = paste(
    "position:fixed; bottom:28px; left:10px;",
    "background:rgba(255,255,255,0.88); padding:5px 10px; border-radius:4px;",
    "box-shadow:0 1px 4px rgba(0,0,0,.2); z-index:900;",
    "font-family:sans-serif; font-size:10px; color:#555; max-width:320px;"
  ),
  HTML(paste0(
    nrow(venues_sf), " geocoded venues (~32% of known venues). ",
    "Northern concentration may reflect geocoding bias. ",
    "Sources: NHGIS, U.S. Census 1980–2000, CEDISH/UTSA."
  ))
)

leaflet_map <- leaflet(options = leafletOptions(zoomControl = TRUE)) %>%

  # ── Tile base layers ────────────────────────────────────────────────────────
  addProviderTiles("OpenStreetMap",      group = "OpenStreetMap") %>%
  addProviderTiles("CartoDB.DarkMatter", group = "Dark Map") %>%

  # ════════════════════════════════════════════════════════════════════════════
  # ECONOMIC LAYERS
  # ════════════════════════════════════════════════════════════════════════════

  addPolygons(
    data = tracts_full, group = "Median Rent 1980",
    fillColor = ~pal_rent_1980(rent_1980), fillOpacity = 0.7,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · $", rent_1980, " (1980)"),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_rent_1980, values = tracts_full$rent_1980,
            title = "Median Rent<br>1980 ($)", opacity = 0.85,
            group = "Median Rent 1980", layerId = "leg_r80") %>%

  addPolygons(
    data = tracts_full, group = "Median Rent 2000",
    fillColor = ~pal_rent_2000(rent_2000), fillOpacity = 0.7,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · $", rent_2000, " (2000)"),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_rent_2000, values = tracts_full$rent_2000,
            title = "Median Rent<br>2000 ($)", opacity = 0.85,
            group = "Median Rent 2000", layerId = "leg_r00") %>%

  addPolygons(
    data = tracts_full, group = "Rent Change 1980→2000",
    fillColor = ~pal_rent_change(rent_change_80_00), fillOpacity = 0.75,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · Δ$",
                    ifelse(is.na(rent_change_80_00), "N/A",
                           format(round(rent_change_80_00), big.mark=","))),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_rent_change,
            values = tracts_full$rent_change_80_00,
            title = "Rent Change<br>1980→2000 ($)<br><small>Red = rising pressure</small>",
            opacity = 0.85, group = "Rent Change 1980→2000",
            layerId = "leg_rchg") %>%

  # ════════════════════════════════════════════════════════════════════════════
  # RACE / ETHNICITY LAYERS
  # ════════════════════════════════════════════════════════════════════════════

  addPolygons(
    data = tracts_full, group = "% Hispanic (2000)",
    fillColor = ~pal_hispanic(pct_hispanic), fillOpacity = 0.7,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · ",
                    ifelse(is.na(pct_hispanic), "N/A",
                           paste0(pct_hispanic, "% Hispanic"))),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_hispanic, values = tracts_full$pct_hispanic,
            title = "% Hispanic<br>(2000)", opacity = 0.85,
            group = "% Hispanic (2000)", layerId = "leg_hisp") %>%

  addPolygons(
    data = tracts_full, group = "% Black Population (2000)",
    fillColor = ~pal_black(pct_black), fillOpacity = 0.7,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · ",
                    ifelse(is.na(pct_black), "N/A",
                           paste0(pct_black, "% Black"))),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_black, values = tracts_full$pct_black,
            title = "% Black Pop<br>(2000)", opacity = 0.85,
            group = "% Black Population (2000)", layerId = "leg_blk") %>%

  # ════════════════════════════════════════════════════════════════════════════
  # DISPLACEMENT VULNERABILITY LAYERS
  # ════════════════════════════════════════════════════════════════════════════

  addPolygons(
    data = tracts_full, group = "% Renter-Occupied (2000)",
    fillColor = ~pal_renter(pct_renter), fillOpacity = 0.7,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · ",
                    ifelse(is.na(pct_renter), "N/A",
                           paste0(pct_renter, "% renter-occupied"))),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_renter, values = tracts_full$pct_renter,
            title = "% Renter-Occ<br>(2000)<br><small>High = more vulnerable</small>",
            opacity = 0.85, group = "% Renter-Occupied (2000)",
            layerId = "leg_rent") %>%

  addPolygons(
    data = tracts_full, group = "Displacement Risk Index",
    fillColor = ~pal_risk(displacement_risk_idx), fillOpacity = 0.72,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · Risk index: ",
                    ifelse(is.na(displacement_risk_idx), "N/A",
                           displacement_risk_idx)),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_risk,
            values = tracts_full$displacement_risk_idx,
            title = "Displacement Risk<br>Index (0–100)<br><small>Rent change × renter %</small>",
            opacity = 0.85, group = "Displacement Risk Index",
            layerId = "leg_risk") %>%

  # ════════════════════════════════════════════════════════════════════════════
  # CULTURAL IMPACT LAYERS
  # ════════════════════════════════════════════════════════════════════════════

  addPolygons(
    data = tracts_full, group = "Venue Churn 1980s→2000s",
    fillColor = ~pal_venue_churn(net_venue_change), fillOpacity = 0.75,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT,
                    " · Lost: ", venues_lost,
                    " · Gained: ", venues_gained,
                    " · Net: ", net_venue_change),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_venue_churn,
            values = tracts_full$net_venue_change,
            title = "Net Venue Change<br>1980s→2000s<br><small>Red = net loss</small>",
            opacity = 0.85, group = "Venue Churn 1980s→2000s",
            layerId = "leg_churn") %>%

  addPolygons(
    data = tracts_full, group = "Event Volume Change 1980s→2000s",
    fillColor = ~pal_event_change(event_change_pct), fillOpacity = 0.72,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT,
                    " · 1980s: ", events_1980s,
                    " events · 2000s: ", events_2000s,
                    " · Change: ",
                    ifelse(is.na(event_change_pct), "N/A",
                           paste0(event_change_pct, "%"))),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_event_change,
            values = tracts_full$event_change_pct,
            title = "Event Volume<br>Change (%)<br>1980s→2000s<br><small>Red = collapse</small>",
            opacity = 0.85, group = "Event Volume Change 1980s→2000s",
            layerId = "leg_evol") %>%

  # ── Dominant Genre 1980s (categorical choropleth) ────────────────────────────
  addPolygons(
    data = tracts_full, group = "Dominant Genre 1980s",
    fillColor = ~pal_dom_genre(genre_1980s), fillOpacity = 0.72,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · ",
                    ifelse(is.na(genre_1980s), "No data", genre_1980s),
                    " (dominant 1980s)"),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_dom_genre,
            values = tracts_full$genre_1980s,
            title = "Dominant Genre<br>1980s (by tract)",
            opacity = 0.85, group = "Dominant Genre 1980s",
            layerId = "leg_g80") %>%

  addPolygons(
    data = tracts_full, group = "Dominant Genre 2000s",
    fillColor = ~pal_dom_genre(genre_2000s), fillOpacity = 0.72,
    color = "white", weight = 0.5, smoothFactor = 1,
    label = ~paste0("Tract ", TRACT, " · ",
                    ifelse(is.na(genre_2000s), "No data", genre_2000s),
                    " (dominant 2000s)",
                    ifelse(isTRUE(genre_shifted), " ⚠ SHIFTED", "")),
    labelOptions = labelOptions(style = list("font-size" = "12px"))
  ) %>%
  addLegend("bottomright", pal = pal_dom_genre,
            values = tracts_full$genre_2000s,
            title = "Dominant Genre<br>2000s (by tract)",
            opacity = 0.85, group = "Dominant Genre 2000s",
            layerId = "leg_g00") %>%

  # ════════════════════════════════════════════════════════════════════════════
  # VENUE POINT LAYERS
  # ════════════════════════════════════════════════════════════════════════════

  addCircleMarkers(
    data = venues_sf, group = "All Venues",
    radius = 6, color = "white", weight = 1,
    fillColor = ~pal_genre(str_extract(top_genres, "^[^,]+")),
    fillOpacity = 0.85, popup = ~popup_html,
    clusterOptions = markerClusterOptions(showCoverageOnHover = FALSE,
                                          maxClusterRadius = 40)
  ) %>%
  addLegend("topright", pal = pal_genre, values = genre_levels,
            title = "Music Genre<br>(All Venues)", opacity = 0.85,
            group = "All Venues") %>%

  addCircleMarkers(data = venues_1970s, group = "Venues 1970s",
    radius = 5, color = "white", weight = 1,
    fillColor = "#e6ab02", fillOpacity = 0.85, popup = ~popup_html) %>%

  addCircleMarkers(data = venues_1980s, group = "Venues 1980s",
    radius = 5, color = "white", weight = 1,
    fillColor = "#1b9e77", fillOpacity = 0.85, popup = ~popup_html) %>%

  addCircleMarkers(data = venues_1990s, group = "Venues 1990s",
    radius = 5, color = "white", weight = 1,
    fillColor = "#d95f02", fillOpacity = 0.85, popup = ~popup_html) %>%

  addCircleMarkers(data = venues_2000s, group = "Venues 2000s",
    radius = 5, color = "white", weight = 1,
    fillColor = "#7570b3", fillOpacity = 0.85, popup = ~popup_html) %>%

  # ── Venue churn point markers ────────────────────────────────────────────────
  addCircleMarkers(
    data = venues_lost_sf, group = "Venues Lost (1980s→2000s)",
    radius = 7, color = "#7b0000", weight = 2,
    fillColor = "#e31a1c", fillOpacity = 0.9,
    popup = ~popup_html
  ) %>%

  addCircleMarkers(
    data = venues_gained_sf, group = "Venues Gained (new in 2000s)",
    radius = 7, color = "#005a00", weight = 2,
    fillColor = "#33a02c", fillOpacity = 0.9,
    popup = ~popup_html
  ) %>%

  # ════════════════════════════════════════════════════════════════════════════
  # LAYER CONTROL
  # ════════════════════════════════════════════════════════════════════════════

  addLayersControl(
    baseGroups    = c("OpenStreetMap", "Dark Map"),
    overlayGroups = c(
      # Economic
      "Median Rent 1980",
      "Median Rent 2000",
      "Rent Change 1980→2000",
      # Race
      "% Hispanic (2000)",
      "% Black Population (2000)",
      # Displacement vulnerability
      "% Renter-Occupied (2000)",
      "Displacement Risk Index",
      # Cultural impact
      "Venue Churn 1980s→2000s",
      "Event Volume Change 1980s→2000s",
      "Dominant Genre 1980s",
      "Dominant Genre 2000s",
      # Venue points
      "All Venues",
      "Venues 1970s",
      "Venues 1980s",
      "Venues 1990s",
      "Venues 2000s",
      "Venues Lost (1980s→2000s)",
      "Venues Gained (new in 2000s)"
    ),
    options = layersControlOptions(collapsed = TRUE)
  ) %>%

  # Default visible: Rent Change + All Venues + Venues Lost
  hideGroup(c(
    "Median Rent 1980",
    "Median Rent 2000",
    "% Hispanic (2000)",
    "% Black Population (2000)",
    "% Renter-Occupied (2000)",
    "Displacement Risk Index",
    "Venue Churn 1980s→2000s",
    "Event Volume Change 1980s→2000s",
    "Dominant Genre 1980s",
    "Dominant Genre 2000s",
    "Venues 1970s",
    "Venues 1980s",
    "Venues 1990s",
    "Venues 2000s",
    "Venues Gained (new in 2000s)"
  )) %>%

  # ── Map controls ─────────────────────────────────────────────────────────────
  setView(lng = -98.4936, lat = 29.4241, zoom = 11) %>%
  addMiniMap(toggleDisplay = TRUE, minimized = TRUE) %>%
  addScaleBar(position = "bottomleft") %>%
  addControl(title_html, position = "topleft") %>%
  addControl(note_html,  position = "bottomleft")

leaflet_map
```

---

# Step 7 — Console Summary

```{r summary}
cat("=== Sounds of San Anto — Map Summary ===\n\n")

cat("── Events ──────────────────────────────\n")
cat("Total in dataset:   ", nrow(events_raw), "\n")
cat("Geocoded (usable):  ", nrow(events),     "\n")
cat("Missing genre data: ", sum(events$macro_genre == "Unknown/Other"), "\n\n")

cat("── Venues ──────────────────────────────\n")
cat("Plotted (All Venues):    ", nrow(venues_sf), "\n")
cat("Zero matched events:     ", sum(venues_sf$n_events == 0), "\n")
cat("Lost (1980s → 2000s):    ", nrow(venues_lost_sf), "\n")
cat("Gained (new in 2000s):   ", nrow(venues_gained_sf), "\n\n")

cat("── Tracts (278 total) ──────────────────\n")
cat("With rent_1980:        ", sum(!is.na(tracts_full$rent_1980)), "\n")
cat("With rent_2000:        ", sum(!is.na(tracts_full$rent_2000)), "\n")
cat("With income_2000:      ", sum(!is.na(tracts_full$income_2000)), "\n")
cat("With Hispanic %:       ", sum(!is.na(tracts_full$pct_hispanic)), "\n")
cat("With Black %:          ", sum(!is.na(tracts_full$pct_black)), "\n")
cat("With renter %:         ", sum(!is.na(tracts_full$pct_renter)),            "\n")
cat("With risk index:       ", sum(!is.na(tracts_full$displacement_risk_idx)), "\n")
cat("Lost ≥1 venue:         ", sum(tracts_full$venues_lost > 0), "\n")
cat("Event volume decline:  ",
    sum(tracts_full$event_change_abs < 0, na.rm = TRUE), "\n")
cat("Genre shifted:         ",
    sum(tracts_full$genre_shifted, na.rm = TRUE), "\n\n")

cat("── Macro-genre breakdown (events) ──────\n")
print(sort(table(events$macro_genre), decreasing = TRUE))
```

---

# Step 8 — Save Map

```{r save-map}
saveWidget(
  leaflet_map,
  file          = "sounds_san_anto_map.html",
  selfcontained = TRUE,
  title         = "Sounds of San Anto: Music Venues & Neighborhood Change"
)
cat("Saved: sounds_san_anto_map.html\n")
```

---

# Findings & Interpretation

**Venue geography and rent pressure**: Venues on the north side of San Antonio
largely occupy tracts with the steepest rent increases 1980–2000 — both because
those corridors were commercially active and because the dataset has a documented
northside collector bias.

**Hispanic and Black geography**: The % Hispanic and % Black layers reveal
distinct spatial patterns. Hispanic population concentrates on the south and
west sides; Black population concentrates on the eastside. Overlaying these with
the Rent Change layer shows that many majority-minority tracts experienced
significant rent pressure with minimal venue investment — consistent with
cultural underinvestment in gentrifying minority neighborhoods.

**Renter vulnerability**: Tracts with high renter-occupied rates (orange layer)
and high rent increases are the most displacement-vulnerable. The southside and
eastside show this combination clearly — the residents with the least ability to
absorb rising housing costs.

**Displacement Risk Index**: The composite layer (rent change × renter %)
synthesizes economic pressure and residential vulnerability into a single score.
High-scoring tracts (deep red/orange) are the neighborhoods where gentrification
most directly threatened the resident population — and where the venue loss and
event volume collapse layers should show the greatest cultural erosion.

**Venue churn and event volume collapse**: The "Venues Lost" red markers and
the Venue Churn choropleth together show where the music economy contracted
most sharply. Tracts with high rent increases tend to show either venue
displacement (venues pushed to cheaper areas) or venue gentrification (new
venues catering to incoming higher-income residents).

**Genre displacement**: The Dominant Genre 1980s vs. 2000s layers make
cultural displacement legible spatially — tracts where Tejano/Latin was
dominant in the 1980s but shifted to Rock or Country by the 2000s map onto
the same corridors experiencing demographic change and rent pressure.

**Limitations**: 312 of ~974 known venues (~32%) are geocoded. Tract boundary
changes between 1980–2000 limit perfect cross-year comparisons. Rent and
income figures are nominal (not inflation-adjusted). Genre displacement
analysis is limited to the ~38% of events that have genre data.
